Benchmarking R packages for calculation of persistent homology

### 2020-04-05 tdabench-paper/tdabench.Rmd
updated citation and title
revised abstract

---
title: Comparing R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  Several persistent homology software libraries have been implemented in R. The Dionysus (TDA), GUDHI (TDA), and Ripser C++ libraries (TDAstats) provide Vietoris-Rips complex functionality. GUDHI  (TDA package) also provides alpha complex functionality. These software represent powerful analysis tools but are limited computationally. These limitations and guidelines for which persistent homology libraries to use have not been specified in R. We benchmarked the all libraries on a High-Powered Cluster for runtime. We also compared alpha and Vietoris-Rips for memory burden. Alpha complexes run significantly faster than all Vietoris-Rips complex libraries for all data sets for data dimension less than or equal to three. Alpha complexes were not computable for data dimension greater than 3. TDAstats had the fastest runtimes for all Vietoris-Rips complexes. We recommend performing persistent homology with GUDHI alpha complexes for data dimension less than or equal to three. For higher dimension, we recommend TDAstats Vietoris-Rips Complex. 
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

>>>>

---
title: Benchmarking R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  Several persistent homology software libraries have been implemented in R. Specifically, the Dionysus, GUDHI, and Ripser libraries have been wrapped in R by the TDA and TDAstats packages on CRAN. They provide Vietoris-Rips and alpha complex functionality. These software represent powerful analysis tools that are currently computationally expensive and, to our knowledge, have not been formally benchmarked in R. Here, we compare the 2 R packages and the 3 underlying libraries for runtime and memory. We find that datasets with less than 3 dimensions can be evaluated with persistent homology fastest by the GUDHI library in the TDA package. For larger dimensional datasets, the Ripser library in the TDAstats package is best. Analysis of runtime and memory growth is also included in this report.
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---
